---
title: "Course Project 1"
output: html_document
---

This document answers all questions assigned in the first course project for the Reproducible Research Coursera course

**Loading and preprocessing the data**

This code reads in the data:

```{r, echo=TRUE}
activity<-read.csv("activity.csv")
```

**What is the mean total number of steps taken per day?**

This code calculates the total number of steps per day:

```{r, echo=TRUE}
total.steps<-tapply(activity$steps, activity$date, sum, na.rm=TRUE)
total.steps<-as.data.frame(total.steps)
total.steps
```

This code creates a histogram of the total number of steps per day:

```{r, echo=TRUE}
hist(total.steps$total.steps,xlab="Number of Steps",ylab="Number of Days",
     main="Totals Steps Per Day",breaks=10,col="green")
```

This code calculates the mean and median of the total number of steps per day:

```{r, echo=TRUE}
#Mean
mean(total.steps$total.steps)
#Median
median(total.steps$total.steps)
```

**What is the average daily activity pattern?**

This codes constructs a plot of the average number of steps taken during each interval:

```{r,echo=TRUE}
mean.steps.interval<-tapply(activity$steps, activity$interval, mean, na.rm=TRUE)
activity1<-data.frame(activity$interval[1:288],mean.steps.interval)
plot(activity1,xlab="Interval",ylab="Mean Number of Steps",
     main="Average Number of Steps Taken in Each 5 Minute Interval")
```

This code finds the interval during which the maximum average number of steps occurs:

```{r, echo=TRUE}
#Output contains interval followed by position in data frame
which.max(mean.steps.interval)
```

**Imputing missing values**

This code calculates the number of missing values in the dataset:

```{r, echo=TRUE}
sum(is.na(activity$steps))
```

This code fills in the missing values for each day with the average number of steps taken that day:

```{r, echo=TRUE}
mean.per.day<-tapply(activity$steps, activity$date, mean)
mean.per.day[is.na(mean.per.day)]<-0
replace<-rep(mean.per.day, each=288)
activity$steps[is.na(activity$steps)] <- replace[is.na(activity$steps)]
```

This code creates a histogram and calculates the mean and median of the total number of steps per day adjusted for the missing values:

```{r, echo=TRUE}
total.steps<-tapply(activity$steps, activity$date, sum, na.rm=TRUE)
total.steps<-as.data.frame(total.steps)
hist(total.steps$total.steps,xlab="Number of Steps",ylab="Number of Days",
     main="Totals Steps Per Day",breaks=10,col="green")
#Mean
mean(total.steps$total.steps)
#Median
median(total.steps$total.steps)
```

This method of imputing the missing data has no effect of the estimates made earlier in the assignment.

**Are there differences in activity patterns between weekdays and weekends?**

This code assigns each day of activity as either a weekday or weekend:

```{r, echo=TRUE}
activity$day<-weekdays(as.Date(activity$date))
activity$day.type<-ifelse(activity$day == "Saturday"|activity$day == "Sunday",
                          "Weekend", "Weekday")
```

This codes constructs a plot of the average number of steps taken during each interval, separated by weekday and weekend:

```{r, echo=TRUE}
par(mfrow=c(2,1))
activity.weekday<-subset(activity,activity$day.type=="Weekday")
mean.steps.weekday<-tapply(activity.weekday$steps, activity.weekday$interval, mean)
activity1.weekday<-data.frame(activity$interval[1:288],mean.steps.weekday)
plot(activity1.weekday,xlab="Interval",ylab="Mean Number of Steps",
     main="Average Number of Steps Taken in Each 5 Minute Interval (Weekdays)")
activity.weekend<-subset(activity,activity$day.type=="Weekend")
mean.steps.weekend<-tapply(activity.weekend$steps, activity.weekend$interval, mean)
activity1.weekend<-data.frame(activity$interval[1:288],mean.steps.weekend)
plot(activity1.weekend,xlab="Interval",ylab="Mean Number of Steps",
     main="Average Number of Steps Taken in Each 5 Minute Interval (Weekends)")
```